---
title: "Lecture 7: Examples"
subtitle: "Mathematics and Statistics BS1515"
author: "Theodoros Arvanitopoulos"
date: "`r Sys.Date()`"
output:
  slidy_presentation:
    font_adjustment: 0
    footer: "Arvanitopoulos T - Lecture 7: Examples"
css: cardiff.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!requireNamespace("gtools", quietly = TRUE)) {
  install.packages("gtools")
}
library(gtools)
```



## Example - Permutation 

### Problem overview

Suppose that two letters are to be selected from A, B, C, D and arranged in order. \

➡️ How many permutations are possible?

---

## Example - Permutation 

### Problem overview

Suppose that two letters are to be selected from A, B, C, D and arranged in order. \

➡️ How many permutations are possible?

### Problem Solution 

We are choosing **2 letters** from a total of **4**, where **order matters**.

$$
P_2^4 = \frac{4!}{(4 - 2)!} = \frac{4!}{2!} = \frac{24}{2} = 12
$$

The possible permutations are:  

AB, AC, AD, BA, BC, BD, CA, CB, CD, DA, DB, DC


## R Example – Permutations

```{r permutations-example}
# Generate all ordered permutations of 2 letters from 4
letters <- c("A", "B", "C", "D")
perms <- gtools::permutations(n = 4, r = 2, v = letters)
head(perms)  # Show first few
nrow(perms)  # Should return 12
```



## Example - Combination 

### Problem overview

Now suppose the same two letters are to be selected from A, B, C, D, but **order does not matter**. \

➡️ How many combinations are possible?


---


## Example - Combination 

### Problem overview

Now suppose the same two letters are to be selected from A, B, C, D, but **order does not matter**. \

➡️ How many combinations are possible?

### Solution

$$
C_2^4 = \frac{4!}{2!(4 - 2)!} = \frac{24}{2! \cdot 2!} = \frac{24}{4} = 6
$$
The possible combinations are:  

AB (same as BA)
AC (same as CA)
AD (same as DA)
BC (same as CB)
BD (same as DB)
CD (same as DC)


## R Example – Combinations

```{r combinations-example}
# Generate all combinations of 2 letters from 4
combs <- combn(letters, 2)
combs
ncol(combs)  # Should return 6
```


## 🧑‍💼 Officer Election (Permutation)

### Problem overview

A club has 8 members. They want to elect a **president**, a **secretary**, and a **treasurer**. Each position must be held by a different person.   \

➡️ How many ways can these positions be filled?

---


## 🧑‍💼 Officer Election (Permutation)

### Problem overview

A club has 8 members. They want to elect a **president**, a **secretary**, and a **treasurer**. Each position must be held by a different person.   \

➡️ How many ways can these positions be filled?

### Solution

This is a **permutation problem** since  order **matters** (president ≠ treasurer, etc).

We are selecting and arranging 3 people out of 8:

$$
P_3^8 = \frac{8!}{(8 - 3)!} = \frac{8!}{5!} = 336
$$


## R Example – Officer Election

```{r officer-election}
# Using factorials directly
factorial(8) / factorial(8 - 3)

# Alternative approach using gtools
positions <- gtools::permutations(n = 8, r = 3)
nrow(positions)  # Should also be 336
```



## Probability Rules – Complement

The probability that an event **does not occur** is $P(\bar{A}) = 1 - P(A)$

➡️  Example: If the probability that a student passes an exam is 0.8, then the probability they fail is:

```{r complement-rule}
P_pass <- 0.8
P_fail <- 1 - P_pass
P_fail
```



## Probability Rules – Addition

For two events $A$ and $B$ $P(A \cup B) = P(A) + P(B) - P(A \cap B)$

➡️  Example: P(A) = 0.6, P(B) = 0.5, P(A and B) = 0.3

```{r addition-rule}
P_A <- 0.6
P_B <- 0.5
P_A_and_B <- 0.3
P_A_or_B <- P_A + P_B - P_A_and_B
P_A_or_B
```



## Probability Rules – Conditional Probability

➡️  Example: P(A and B) = 0.25, P(B) = 0.5. What is P(A given B)?

$P(A \mid B) = \frac{P(A \cap B)}{P(B)}$

```{r conditional-prob}
P_A_and_B <- 0.25
P_B <- 0.5
P_A_given_B <- P_A_and_B / P_B
P_A_given_B
```



## Bayes' Theorem - Medical Screening Example

### Problem overview

Imagine a rare disease affects 1% of the population.A diagnostic test exists for this disease:\

- If a person **has the disease**, the test gives a **positive** result 95% of the time (true positive). \
- If a person **does not have the disease**, the test gives a **false positive** 10% of the time. \

➡️  If a patient tests **positive**, what is the **probability they actually have the disease**?

---

## Bayes' Theorem - Medical Screening Example

### Problem overview

Imagine a rare disease affects 1% of the population.A diagnostic test exists for this disease:\

- If a person **has the disease**, the test gives a **positive** result 95% of the time (true positive). \
- If a person **does not have the disease**, the test gives a **false positive** 10% of the time. \

➡️ If a patient tests **positive**, what is the **probability they actually have the disease**?


### Step 1: Define the Given Probabilities

Let:\
- $P(D) = 0.01$: probability of having the disease \
- $P(\sim D) = 0.9$: probability of **not** having the disease \
- $P(Pos \mid D) = 0.95$: probability of testing positive **given disease** \
- $P(Pos \mid \sim D) = 0.10$: probability of testing positive **given no disease** \



## Bayes' Theorem - Problem solution

### Step 2: Apply Bayes' Theorem


The total probability of testing positive:

$$
P(Pos) = P(Pos \mid D) \cdot P(D) + P(Pos \mid \sim D) \cdot P(\sim D)= \\
(0.95)(0.01) + (0.10)(0.99) \\
=  0.0095 + 0.099 = 0.1085
$$
Apply Bayes’ theorem:

$$
P(D \mid Pos) = \frac{P(Pos \mid D) \cdot P(D)}{P(Pos)}= \frac{(0.95)(0.01)}{0.1085}= \frac{0.0095}{0.1085} \approx 0.0876
$$




Even though the test is quite accurate, a **positive result only means an 8.76% chance** the person actually has the disease — due to the **low base rate**.

This is a classic case where even with a good test, the **post-test probability** is much lower than intuition may suggest, due to the rarity of the disease.

This highlights the **importance of prior probabilities** in understanding test results — a classic insight from Bayes' Theorem.



## Bayes' Theorem - Problem solution using R

```{r, message=FALSE}
# Known probabilities
P_Disease <- 0.01                         # P(D)
P_Positive_given_Disease <- 0.95          # P(Pos | D)
P_Positive_given_NoDisease <- 0.10        # P(Pos | ~D)

# Complementary probability
P_NoDisease <- 1 - P_Disease              # P(~D)

# Total probability of testing positive
P_Positive <- P_Positive_given_Disease * P_Disease +
              P_Positive_given_NoDisease * P_NoDisease

# Apply Bayes' Theorem
P_Disease_given_Positive <- (P_Positive_given_Disease * P_Disease) / P_Positive

# Output result
P_Disease_given_Positive

```



## Binomial Distribution - Example

### Problem overview

- The number of trials $n = 10$ \
- The probability of success in each trial $p = 0.4$ \
  
➡️ What is the probability of 1 success in 10 observations if the probability of success is 0.4?

---

## Binomial Distribution - Example

### Problem overview

- The number of trials $n = 10$ \
- The probability of success in each trial $p = 0.4$ \
  
➡️ What is the probability of 1 success in 10 observations if the probability of success is 0.4?


### Problem solution

We model this situation using the **Binomial Distribution**, which gives the probability of $x$ successes in $n$ independent trials:

$$
P(X = x\mid n, p) = \binom{n}{x} p^x (1 - p)^{n - x}
$$


In our case:
$$
P(X = 1) = \binom{10}{1} (0.4)^1 (0.6)^9=\frac{10!}{1!(10-1)!}(0.4)^1 (0.6)^9= 0.0403
$$



## Binomial Distribution - Problem solution using R

```{r, echo=TRUE}
# Compute probability of exactly 1 success
# P(x = 1 | n = 10, p = 0.4)
dbinom(1, size = 10, prob = 0.4)

# Create table of probabilities for all possible outcomes (0 to 10)
x <- 0:10
prob <- dbinom(x, size = 10, prob = 0.4)
data.frame(Successes = x, Probability = round(prob, 4))
```



## Using a binomial table (n = 10) {.smaller}

### Binomial table (n = 10)

This is a Binomial table for $n = 10$ and a range of $p$ values.  \ 
Each **row** corresponds to a value of $x$ (number of 'successes'), and each **column** corresponds to a value of the success probability $p$.  \

- **Example 1:** $n = 10, x = 1, p = 0.4$ → expected probability **0.0403**.  \
- **Example 2:** $n = 10, x = 8, p = 0.45$ → expected probability **0.0229**.


```{r build-binom-table, echo=FALSE}
# Ensure packages
if (!requireNamespace("kableExtra", quietly = TRUE)) {
  install.packages("kableExtra")
}
library(kableExtra)

n <- 10
x_vals <- 0:10
p_vals <- c(0.20, 0.25, 0.3, 0.35, 0.4, 0.45, 0.5, 0.55)

# Matrix of probabilities
prob_mat <- sapply(p_vals, function(p) dbinom(x_vals, size = n, prob = p))

# Round for display
prob_df <- as.data.frame(round(prob_mat, 4))
colnames(prob_df) <- paste0("p=", p_vals)
prob_df <- cbind(x = x_vals, prob_df)

prob_df
```

### Check the Highlighted Values Programmatically 

```{r check-highlighted}
P_x1_p040 <- dbinom(1, size = 10, prob = 0.40)
P_x8_p045 <- dbinom(8, size = 10, prob = 0.45)
c("P(X=1|n=10,p=0.40)" = P_x1_p040,
  "P(X=8|n=10,p=0.45)" = P_x8_p045)
```



## Binomial Distribution — Example 

### Problem overview

The number of trials becomes $n = 12$, with probability of success in each trial $p = 0.45$ \
  
➡️ What is the expected success, and what the variance and standard deviation of success.
  
 \(X \sim \text{Binomial}(n=12,\ p=0.45)\) 
 
\[
\mathbb E[X]=np,\qquad \mathrm{Var}(X)=np(1-p),\qquad \mathrm{SD}(X)=\sqrt{np(1-p)}.
\]

\[
\mathbb E[X]=12\cdot0.45=5.4,\quad 
\mathrm{Var}(X)=12\cdot0.45\cdot0.55=2.97,\quad
\mathrm{SD}(X)=\sqrt{2.97}\approx 1.723.
\]

```{r, echo=TRUE}
# Parameters
n  <- 12
p  <- 0.45

# Binomial mean, variance, sd
E_X   <- n * p
Var_X <- n * p * (1 - p)
SD_X  <- sqrt(Var_X)

c(Mean = E_X, Variance = Var_X, SD = SD_X)
```
