---
title: "Lecture 9: Examples"
subtitle: "Mathematics and Statistics BS1515"
author: "Theodoros Arvanitopoulos"
date: "`r Sys.Date()`"
output:
  slidy_presentation:
    font_adjustment: 0
    footer: "Arvanitopoulos T - Lecture 9: Examples"
css: cardiff.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( warning=FALSE, message=FALSE)
library(ggplot2)
```


## Example 1 - Confidence intervals when $\sigma$ known {.smaller}

### Problem overview 

A sample of 11 circuits from a large normal population has a mean resistance of 2.20 ohms. We know from past testing that the population standard deviation is 0.35 ohms.  \

Determine a 95% confidence interval for the true mean resistance of the population.

---

## Example 1 - Confidence intervals when $\sigma$ known {.smaller}

### Problem overview 

A sample of 11 circuits from a large normal population has a mean resistance of 2.20 ohms. We know from past testing that the population standard deviation is 0.35 ohms.  \

Determine a 95% confidence interval for the true mean resistance of the population.

### Problem solution

We use the formula for the confidence interval of the population mean when $\sigma$ is known

$$
\bar{x} \pm \frac{\sigma}{\sqrt{n}} \\
= 2.20 \pm 1.96 (0.35 / \sqrt{11}) \\
= 2.20 \pm 0.2068 \\
$$

So, the interval is 
$$
1.9932 < \mu < 2.4068
$$

---

## Example 1 - Confidence intervals when $\sigma$ known {.smaller}

### Problem overview 

A sample of 11 circuits from a large normal population has a mean resistance of 2.20 ohms. We know from past testing that the population standard deviation is 0.35 ohms.  \

Determine a 95% confidence interval for the true mean resistance of the population.

### Problem solution

We use the formula for the confidence interval of the population mean when $\sigma$ is known

$$
\bar{x} \pm \frac{\sigma}{\sqrt{n}} \\
= 2.20 \pm 1.96 (0.35 / \sqrt{11}) \\
= 2.20 \pm 0.2068 \\
$$

So, the interval is 
$$
1.9932 < \mu < 2.4068
$$

### Interpretation  

We are **95% confident** that the true mean resistance of the population lies between 1.9932 ohms and 2.4068 ohms.

This does not mean the true mean is definitely in this range. It means that if we repeated this sampling method many times, 95% of such intervals would contain the true population mean.



## Example 1 - Problem solution using R {.smaller}

```{r ci-known-sigma, message=FALSE}
# Known values from the problem
x_bar <- 2.20        # Sample mean
sigma <- 0.35        # Population standard deviation
n <- 11              # Sample size
z_critical <- 1.96   # Z value for 95% confidence level

# Standard error
se <- sigma / sqrt(n)

# Margin of error
margin_error <- z_critical * se

# Confidence interval
ci_lower <- x_bar - margin_error
ci_upper <- x_bar + margin_error

# Output the interval with explanation
cat("95% Confidence Interval:\n")
cat(sprintf("Lower bound: %.4f\n", ci_lower))
cat(sprintf("Upper bound: %.4f\n", ci_upper))
```



## Example 2 - Confidence Interval when $\sigma$ unknown {.smaller}

### Problem overview

A random sample of 16 gas grills has a mean ignition time of 0.51 seconds and a sample standard deviation of 0.08 seconds.Assume ignition time is approximately normally distributed. 

Construct a 95% confidence interval for the true mean ignition time.

---

## Example 2 - Confidence Interval when $\sigma$ unknown {.smaller}

### Problem overview

A random sample of 16 gas grills has a mean ignition time of 0.51 seconds and a sample standard deviation of 0.08 seconds.Assume ignition time is approximately normally distributed. 

Construct a 95% confidence interval for the true mean ignition time.

### Problem solution

Since $\sigma$ is not known, we given this is approximately normally distributed, we use the t-distribution. Degrees of freedom: $df = n - 1 = 15$ and thus $t_{df=15, \ \alpha/2} = 2.131$ from t-table

We use the formula for the confidence interval : \

$$
\bar{x} - t_{15, \ \alpha/2} \cdot \frac{S}{\sqrt{n}} < \mu < \bar{x} + t_{15, \ \alpha/2} \cdot \frac{S}{\sqrt{n}} \\
0.51-(2.131) \frac{0.08}{4} < \mu < 0.51+(2.131) \frac{0.08}{4} \\
$$

So, the interval is 
$$
0.4673 < \mu < 0.5526
$$

---

## Example 2 - Confidence Interval when $\sigma$ unknown {.smaller}

### Problem overview

A random sample of 16 gas grills has a mean ignition time of 0.51 seconds and a sample standard deviation of 0.08 seconds.Assume ignition time is approximately normally distributed. 

Construct a 95% confidence interval for the true mean ignition time.

### Problem solution

Since $\sigma$ is not known, we given this is approximately normally distributed, we use the t-distribution. Degrees of freedom: $df = n - 1 = 15$ and thus $t_{df=15, \ \alpha/2} = 2.131$ from t-table

We use the formula for the confidence interval : \

$$
\bar{x} - t_{15, \ \alpha/2} \cdot \frac{S}{\sqrt{n}} < \mu < \bar{x} + t_{15, \ \alpha/2} \cdot \frac{S}{\sqrt{n}} \\
0.51-(2.131) \frac{0.08}{4} < \mu < 0.51+(2.131) \frac{0.08}{4} \\
$$

So, the interval is 
$$
0.4673 < \mu < 0.5526
$$

### Interpretation

We are 95% confident that the true mean ignition time is between 0.4673 seconds and 0.5526 seconds. \

This means that 95% of such intervals, computed in the same way from random samples of size 16, would contain the true population mean.



## Example 2 - Problem solution using R {.smaller}



```{r}
# Sample statistics
x_bar <- 0.51         # Sample mean
s <- 0.08             # Sample standard deviation
n <- 16               # Sample size
df <- n - 1           # Degrees of freedom
t_critical <- 2.131   # From t-table for 95% CI and df = 15

# Standard error
se <- s / sqrt(n)

# Margin of error
error_margin <- t_critical * se

# Confidence interval bounds
lower <- x_bar - error_margin
upper <- x_bar + error_margin

# Print formatted result
cat("95% Confidence Interval (t-distribution):\n")
cat(sprintf("Lower bound: %.4f\n", lower))
cat(sprintf("Upper bound: %.4f\n", upper))
``` 



## Example 3 - Confidence Interval for a Population Proportion {.smaller}

### Problem Overview

A random sample of 100 people reveals that 25 are left-handed. 

We are asked to construct a 95% confidence interval for the true proportion of left-handers in the population.

---

## Example 3 - Confidence Interval for a Population Proportion {.smaller}

### Problem Overview

A random sample of 100 people reveals that 25 are left-handed. 

We are asked to construct a 95% confidence interval for the true proportion of left-handers in the population.

### Problem solution

We use the formula for the confidence interval for population proportions:
$$
\hat{p} - z_{\alpha/2} \frac{\sqrt{\hat{p}(1-\hat{p})}}{n} < P < \hat{p} + z_{\alpha/2} \frac{\sqrt{\hat{p}(1-\hat{p})}}{n} \\
(or \  \hat{p} \pm z_{\alpha/2} \frac{\sqrt{\hat{p}(1-\hat{p})}}{n})  \\
25/100 - 1.96 \sqrt{ \frac{0.25(0.75)}{100}} < P < 25/100 + 1.96 \sqrt{ \frac{0.25(0.75)}{100}}  \\
0.1651 < P < 0.3349
$$

---

## Example 3 - Confidence Interval for a Population Proportion {.smaller}

### Problem Overview

A random sample of 100 people reveals that 25 are left-handed. 

We are asked to construct a 95% confidence interval for the true proportion of left-handers in the population.

### Problem solution

We use the formula for the confidence interval for population proportions:
$$
\hat{p} - z_{\alpha/2} \frac{\sqrt{\hat{p}(1-\hat{p})}}{n} < P < \hat{p} + z_{\alpha/2} \frac{\sqrt{\hat{p}(1-\hat{p})}}{n} \\
(or \  \hat{p} \pm z_{\alpha/2} \frac{\sqrt{\hat{p}(1-\hat{p})}}{n})  \\
25/100 - 1.96 \sqrt{ \frac{0.25(0.75)}{100}} < P < 25/100 + 1.96 \sqrt{ \frac{0.25(0.75)}{100}}  \\
0.1651 < P < 0.3349
$$


### Interpretation

We are 95% confident that the true proportion of left-handers in the population lies between  16.51% and 33.49%.

That is, if we took many samples of 100 people and calculated confidence intervals in the same way, about 95% of them would contain the true population proportion.



## Example 3 - Problem solution using R {.smaller}


```{r}
# Given values
x <- 25                # Number of left-handers
n <- 100               # Sample size
phat <- x / n          # Sample proportion
z <- 1.96              # Z-value for 95% confidence level

# Standard error
se <- sqrt(phat * (1 - phat) / n)

# Margin of error
me <- z * se

# Confidence interval
lower <- round(phat - me, 4)
upper <- round(phat + me, 4)

# Output results
cat("95% Confidence Interval for Population Proportion:\n")
cat(sprintf("Lower bound: %.4f\n", lower))
cat(sprintf("Upper bound: %.4f\n", upper))
```




